---
title: "hypothesis_testing.rmd"
author: "Nikola Baci & Jake Hofman"
date: "June 13, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(ggplot2)
theme_set(theme_bw())
set.seed(42)

```
# Estimating a proportion
## Point estimate and sampling distribution

Experiment: Flip a biased coin 100 times
Repeat the experiment 100,000 times. Each time record the mean of the
experiment in the variable p_hat.
Finally, compare the given bias (p) with the mean of all the means in the p_hat.

Conclusion: 

```{r}
estimate_coin_bais <- function(n, p){
  mean(rbinom(n, 1, p))
}

n <- 100
p <- 0.3
p_hat <- replicate(1e5, estimate_coin_bais(n, p))

# Plot the sampling distribution
ggplot(data = data.frame(p_hat))+
  geom_histogram(mapping = aes(x = p_hat), binwidth = 0.01)+
  geom_vline(xintercept=p)+
  geom_vline(xintercept=mean(p_hat), linetype=2, color = "red")
  
print(paste("n =", n, "p = ", p, "p_hat_bar = ", mean(p_hat)))

plot_data <- data.frame()
for(n in c(100, 200, 400, 800)){
  tmp <- data.frame(n = n, p_hat = replicate(1e5, estimate_coin_bias(n, p)))
  plot_data <- rbind(plot_data, tmp)
}

ggplot(data = plot_data)+
  geom_histogram(mapping = aes(x = p_hat), binwidth = 0.01)+
  facet_wrap(~ n, nrow = 1)

se_data <- plot_data %>%
  group_by(n) %>%
  summarize(se  = sd(p_hat))

ggplot(data = se_data)+
  geom_point(mapping = aes(x = n, y = se))+
  stat_function(fun= function(n) {sqrt(p * (1-p) / n)}, linetype = 2)
       
```

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
