---
title: "Movielens"
date: '`r Sys.time()`'
output:
  html_document:
    #code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
library(scales)
library(tidyverse)
library(knitr)
library("zoo")
# set plot theme
theme_set(theme_bw())
```

# Load and preview data

Read data from the `ratings.csv` file

```{r load-data}
ratings <- read_csv('ratings.csv',
                    col_names = c('user_id','movie_id','rating','timestamp'))
```

Loaded `r format(object.size(ratings), units="Mb")` of ratings data, containing `r format(nrow(ratings), big.mark = ",")` ratings. Here's a preview:

```{r preview-data}
head(ratings) %>% kable()
```

# Summary statistics

```{r dist-ratings}
# plot the distribution of rating values https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=26
ratings %>%
  ggplot()+
  geom_bar(aes(x = rating))+
  xlab("Rating") +
  ylab("Number of ratings") +
  scale_y_continuous(labels = comma)

```

#------------------------------------------------------------------------------#
## Per-movie stats

```{r aggregate-by-movie}
# aggregate ratings by movie, computing mean and number of ratings
# hint: use the n() function for easy counting within a group

movie_dist <- ratings %>%
  group_by(movie_id) %>%
  summarize(count = n(),
            average_rating = mean(rating))

# Sanity check: this should give the fist 6 ratings for movie with id 122.
# The count is 6 and the average is 3.5.
head(filter(ratings, movie_id == 122)) %>% 
  group_by(movie_id) %>%
  summarize(count = n(),
            average_rating = sum(rating) / count) %>% View()
```
#------------------------------------------------------------------------------#

```{r dist-movie-popularity}
# plot distribution of movie popularity (= number of ratings the movie received)
# hint: try scale_x_log10() for a logarithmic x axis

movie_dist %>%
  ggplot()+
  geom_histogram(aes(x = count))+
  scale_x_log10(labels = comma) +
  xlab("Number of Ratings per Movie") +
  ylab("Number of Movies")
  

```

```{r dist-mean-ratings-by-movie}
# plot distribution of mean ratings by movie https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=28
# hint: try geom_histogram and geom_density

movie_dist %>%
  ggplot() +
  geom_density(aes(x = average_rating), fill = "black")

movie_dist %>%
  ggplot() +
  geom_histogram(aes(x = average_rating), binwidth = 0.1)
  
```

```{r cdf-movie-pop}
# rank movies by popularity and compute the cdf, or fraction of movies covered by the top-k moves https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=30
# hint: use dplyr's rank and arrange functions, and the base R sum and cumsum functions
# store the result in a new data frame so you can use it in creating figure 2 from the paper below

# plot the CDF of movie popularity

total_rank <- length(movie_dist$movie_id)

ranked_movies <- movie_dist %>%
  arrange(desc(count)) %>%
  mutate(rank = c(1:total_rank))

total_reviews_sum <-  sum(ranked_movies$count)

ranked_movies_with_percent <- ranked_movies %>%
  mutate(percent = count / total_reviews_sum)

ranked_movies_cumsum <- ranked_movies_with_percent %>%
  mutate(cum_sum = cumsum(percent))

ranked_movies_cumsum %>%
  ggplot() +
  geom_line(aes(x = rank, y = cum_sum)) +
  xlab("Movie Rank") +
  ylab("CDF") +
  geom_vline(xintercept = 3000, linetype = 2)

```


# Per-user stats

```{r aggregate-by-user}
# aggregate ratings by user, computing mean and number of ratings
user_dist <- ratings %>%
  group_by(user_id) %>%
  summarize(ratings_count = n(), 
            ratings_mean = mean(rating))
```

```{r dist-user-activity}
# plot distribution of user activity (= number of ratings the user made)
# hint: try a log scale here

user_dist %>%
  ggplot() +
  geom_histogram(aes(x = ratings_count), binwidth = 0.05)+
  scale_x_log10(labels = comma) +
  xlab("Number of Ratings") +
  ylab("Number of Users")
```

# Anatomy of the long tail

```{r long-tail}
# generate the equivalent of figure 2 of this paper:
# https://5harad.com/papers/long_tail.pdf

# Specifically, for the subset of users who rated at least 10 movies,
# produce a plot that shows the fraction of users satisfied (vertical
# axis) as a function of inventory size (horizontal axis). We will
# define "satisfied" as follows: an individual user is satisfied p% of
# the time at inventory of size k if at least p% of the movies they
# rated are contained in the top k most popular movies. As in the
# paper, produce one curve for the 100% user satisfaction level and
# another for 90%---do not, however, bother implementing the null
# model (shown in the dashed lines).
```

First, join the "ranking" table with the "ranked_movies" table computed
previously. This will allow us to have the movie_id, user_id and rank 
in the same table.

```{r}
#df <- ratings
#dataframe <- df %>% left_join(ranked_movies, by="movie_id")
#data <- select(dataframe, c(user_id, movie_id, rank))

data <- ratings %>% left_join(ranked_movies, by="movie_id") %>%
  select(c(user_id, movie_id, rank))

total_users <- nrow(user_dist)

  
```

Second, manipulate the data to get the table that gives the number of users
who are 100% satisfied with a given inventory.

```{r}
user_satisfaction_100 <- data %>%
  group_by(user_id) %>%
  summarize(max_rank = max(rank))

user_satisfaction_100_cumsum <- user_satisfaction_100 %>%
  group_by(max_rank) %>%
  summarize(num_of_users_per_rank_100 = n()) %>% 
  mutate(percent_100 = num_of_users_per_rank_100 / total_users )%>%
  mutate(cum_sum_100 = cumsum(percent_100))

user_satisfaction_100_cumsum %>%
  ggplot()+
  geom_line(aes(x = max_rank, y = cum_sum_100))
```

```{r}
#used to determine the range of row numbers for each user in the "data" table
data <- mutate(data, row = c(1:nrow(data))) 
#used to get the right rank for each 90Th percentile
data <- arrange(data, user_id, rank)

#get the 90th rank movie for each user
user_satisfaction_90 <- data %>%
  group_by(user_id) %>%
  summarize(mini = min(row),
            maxi = max(row),
            row_number = ceiling(quantile(c(mini:maxi), probs = 0.9)),
            data[row_number, "rank"])

#get the cumulative sum for each rank (inventory)
user_satisfaction_90_cumsum <- user_satisfaction_90 %>%
  group_by(rank) %>%
  summarize(num_of_users_per_rank_90 = n()) %>%
  mutate(percent_90 = num_of_users_per_rank_90 / total_users)%>%
  mutate(cum_sum_90 = cumsum(percent_90))

user_satisfaction_90_cumsum %>% 
  ggplot() +
  geom_line(aes(x = rank, y = cum_sum_90))
```


```{r}
final_table <- user_satisfaction_90_cumsum %>% 
  full_join(user_satisfaction_100_cumsum, by = c("rank" = "max_rank"))

#full_join seems to put all rows from "user_satisfaction_90_cumsum" at the 
#end of the table if the row contains a NA value (this case cum_sum_90)
final_table <- arrange(final_table, rank)

final_table <- na.locf(final_table, formLast = T)

final_table %>%
  ggplot(aes(x = rank)) +
  geom_line(aes(y = cum_sum_100)) +
  geom_line(aes(y = cum_sum_90)) +
  xlab("Inventory") +
  ylab("Percent of Users Satisfied")
```
